Peer Assignment 1 - Reproduciple Research Course
========================================================

In this assignment we will make use of data from a **personal activity** monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the *number of steps taken in 5 minute intervals each day*.

## The Dataset

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset. 

The dataset can be downloaded from [Activity monitoring data [52K]](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The variables included in this dataset are:
- **steps:** Number of steps taking in a 5-minute interval (missing values are coded as **NA**)
- **date:** The date on which the measurement was taken in **YYYY-MM-DD**
format
- **interval:** Identifier for the 5-minute interval in which measurement was taken

### Loading and preprocessing the data
1- We will load the dataset bu using the read.csv function.

```{r echo=TRUE}
adata <- read.csv("activity.csv")
summary(adata)
```

2- Convert the date factor to type Date
```{r echo=TRUE}
adata$date <- as.Date(adata$date, format="%Y-%m-%d")
summary(adata)
```

### Mean and median total number of steps taken per day

1- We will first make a histogram of the total number of steps taken each day.

First we use the function **ddply** of the library **plyr** to compute the total number of steps taken per day:

```{r echo=TRUE}
require(plyr)
stepsPerDay <- ddply(adata, .(date), summarize, totalSteps = sum(steps))
```

then we make a histogram by subdivising the total number of steps in ranges (X-axis) and gives the frequencies (Y-axis) of days in every range.

```{r echo=TRUE}
par(col.main = "red", col.lab = "red", col.axis = "blue", cex.axis=0.8)

hist(stepsPerDay$totalSteps, col = "blue", border="white",
     xlab = "Range of Steps per Day", 
     main = "The total number of steps per day")
```

2- We can now compute the **mean** and **median** of total number of steps per day as follow:
```{r echo=TRUE}
meanStepsPerDay <- mean(stepsPerDay$totalSteps, na.rm=TRUE)
meanStepsPerDay
medianStepsPerDay <- median(stepsPerDay$totalSteps, na.rm=TRUE)
medianStepsPerDay
```
So, the mean of total number of steps per day is **`r as.character(round(meanStepsPerDay,2))`** and the median of total number of steps per day is **`r as.character(round(medianStepsPerDay,2))`**.

### The average daily activity pattern

1- First, we will compute the average number of steps across all days for the 5-minute intervals 

```{r echo=TRUE}
meanStepsPerInterval <- ddply(adata, .(interval), summarize, mean = mean(steps, na.rm=TRUE))
summary(meanStepsPerInterval)
```

We will make a graphical presentation of the average number of steps taken (y-axis) for each the 5-minute interval (x-axis), averaged across all days 

```{r echo=TRUE}
# set the color to titles
par(col.main = "red", col.lab = "red", col.axis = "blue", cex.axis=0.9)

plot(meanStepsPerInterval$interval, meanStepsPerInterval$mean , 
     type="l", xaxt = "n", yaxt = "n", col = "blue", 
     main = "The average number of steps per 5-minute intevals",
     xlab = "5-minute Time Intervals (labeled in chunks of 1 hour)", 
     ylab = "Average of Steps Across Days")

indices <- seq(1, nrow(meanStepsPerInterval), by=12)
xlabels <- sprintf("%02d:%02d", meanStepsPerInterval$interval %/% 100, meanStepsPerInterval$interval %% 100)
axis(1, meanStepsPerInterval$interval[indices], xlabels[indices], las=3)
axis(2,las=1)
```

2- We can compute the 5-minute interval that contains the maximum number of steps on average across all days 

```{r echo=TRUE}
indeMaxInterval <- which.max(meanStepsPerInterval$mean)
maxInterval <- meanStepsPerInterval$interval[indeMaxInterval]
maxInterval
intervalTime <- sprintf("%02d:%02d", maxInterval %/% 100, maxInterval %% 100)
intervalTime
```
So, **`r maxInterval`** is the 5-minute interval in question (correspond to the time `r intervalTime`).

### Imputing missing values

1- We can compute the number of missing data by using the functions **is.na** and **table**. As it is required in the assignment, we have to look after NB values by rows whatever the missing values in the dataset, I will not restrein the test to the steps only. So we can use the following script
```{r echo=TRUE}
nbNA <- table((is.na(adata$steps) | is.na(adata$date) | is.na(adata$interval)))[["TRUE"]]
nbNA
```
So, the number of **NA** in the dataset is **`r nbNA`**.

2- For the missing values I will use the average of the corresponding 5-minute interval to replace it.

3- Create the new dataset by replacing missing values by corresponding 5-minute interval averages across all days. So, after getting the averages of 5-minute intervals, we duplicate them as many as there are days to fill in the missing values of steps by the corresponding average ones. 
```{r echo=TRUE}
newData <- adata
# duplicate the the average of interval to match the length of all days
dupmean <- rep(meanStepsPerInterval$mean, length(table(adata$date)))
isNA <- is.na(adata$steps)
newData$steps[isNA] <- dupmean[isNA]
```

4- Compute the total number of steps per day:

```{r echo=TRUE}
stepsPerDay <- ddply(newData, .(date), summarize, totalSteps = sum(steps))
```

then we make a histogram by subdivising the total number of steps in ranges (X-axis) and gives the frequencies (Y-axis) of days in every range.

```{r echo=TRUE}
par(col.main = "red", col.lab = "red", col.axis = "blue", cex.axis=0.9)
hist(stepsPerDay$totalSteps, col = "blue", border="white",
     xlab = "Range of Steps per Day",   
     main = "The total number of steps per day")
```

The **mean** and **median** of total number of steps per day are computed as follow:
```{r echo=TRUE}
meanStepsPerDay <- mean(stepsPerDay$totalSteps)
meanStepsPerDay
medianStepsPerDay <- median(stepsPerDay$totalSteps)
medianStepsPerDay
```
So, the mean of total number of steps per day is **`r as.character(round(meanStepsPerDay,2))`** and the median of total number of steps per day is **`r as.character(round(medianStepsPerDay,2))`**.

In general, there are not difference in histogram form with almost the same values for the mean and ,edian, but the number of steps is becoming bigger especially around the median.


Now, I will introduce a new data field in the new dataset. first I initialize this filed to be completely to the `weeday` value then I modify the values corresponding to `Saturday` and `Sunday` to the value `weekend`.
```{r echo=TRUE}
newData$week <- "weekday"
isWeekend <-weekdays(as.Date(newData$date)) == "Sunday" | 
            weekdays(as.Date(newData$date)) == "Saturday"
newData$week[isWeekend] <- "weekend"
```

After adding the fied week to the new dataset, we can compute the 5-minute interval average across weekday and weekend as follow:

```{r echo=TRUE}
meanAcrossWeekdays <- ddply(newData, .(week, interval), summarize, mean = mean(steps))
summary(meanStepsPerInterval)
```

Finally, we can plot the 5-minute interval steps averaged across weekdays and weekend by using xyplot function of the package `lattice`

```{r echo=TRUE}
require(lattice) 
meanAcrossWeekdays$week <- as.factor(meanAcrossWeekdays$week)
indices <- seq(1, nrow(meanStepsPerInterval), by=12)

# xlabels contains the times corresponding to 5-minute intervals
# indices contains the indices of values to use for x-axis anntations

xyplot(mean ~ interval | week, meanAcrossWeekdays,
       type = c("l", "l"), layout = c(1, 2), ylab = "Average of Steps",
       xlab = "5-minute Intervals (labeled in chunks of 1 hour)",
       scales = list(col = "blue",
                     x = list(labels = xlabels[indices], rot= c(45), 
                              at = meanStepsPerInterval$interval[indices])                     
                  )
       )
```

We have noticed that:

-In **weekend**, people are far active during all the day comparing to the night where they become gradually inactive in the evening, we also noticed more activity early morning around 8 AM.

-In **weekdays**, people are completeley inactive in the night and they become active more early comparing to the weekend starting from 5:30 AM, At 8 AM starts the top activity period till 9 AM where the activity drops dramatically to become marginal all the day except in the period of lunch (12 - 13 PM), the activity returns at the end of the day especially a the period of return fron work (18-19PM) then it returns gradually to almost inactive around 10 PM.

**Thank you**